### 2025-07-01 paper/sections/method.qmd

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

A key empirical finding of @altmeyer2024faithful was that JEMs perform well on the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals --- the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology. Instead we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with regards to the choice for the $\text{advloss}(\cdot)$ term in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we work with a different distinction between CE and AE than @freiesleben2022intriguing who considers misclassification as the distinguishing feature of adversarial examples. One of the key observations of this work is that we can leverage CEs during training and get AEs essentially for free to reap the aforementioned benefits of adversarial training.

>>>>

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

A key empirical finding of @altmeyer2024faithful was that JEMs perform well on the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals --- the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology. Instead we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:
$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with regards to the choice for the $\text{advloss}(\cdot)$ term in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we work with a different distinction between CE and AE than @freiesleben2022intriguing who considers misclassification as the distinguishing feature of adversarial examples. One of the key observations of this work is that we can leverage CEs during training and get AEs essentially for free to reap the aforementioned benefits of adversarial training.

### 2025-07-01 paper/sections/related.qmd

CEs have also been used to improve models in the natural language processing (NLP) domain. First, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models. The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks. Second, @balashankar2023improving use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.

>>>>

CEs have also been used to improve models in the natural language processing domain. For example, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
